---
title: "Commuting Network Analysis Project"
author: "Zeke Jeske"
date: "May 7, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

source("setup.R")

suppressPackageStartupMessages({
  library(igraph)
  library(tidygraph)
  library(ggraph)
  library(tmap)
  library(ggrepel)
  library(blockmodeling)
  library(knitr)
})
```

This report analyzes Chicago's inter-neighborhood commuting network using
LODES origin-destination data for 2002 and 2022. Each of the city's 77
community areas (CCAs) is a node; the number of workers commuting between any
two CCAs is a directed, weighted edge.

```{r functions, include=FALSE}
# ── Graph construction ────────────────────────────────────────────────────────

create_graph <- function(flows) {
  tbl_graph(edges = flows, nodes = ccas) %>%
    activate(edges) %>%
    mutate(
      from_name = .N()$name[from],
      to_name   = .N()$name[to]
    ) %>%
    activate(nodes) %>%
    mutate(
      in_degree  = centrality_degree(weights = n, mode = "in"),
      out_degree = centrality_degree(weights = n, mode = "out"),
      eigen      = centrality_eigen(weights = n, directed = TRUE, scale = TRUE)
    )
}

# Convenience accessor used by map helpers below
g <- function(year) {
  switch(as.character(year),
    "2002" = g2002,
    "2022" = g2022,
    stop("Only years 2002 and 2022 are available")
  )
}

get_sf <- function(year) {
  g(year) %>%
    activate(nodes) %>%
    as_tibble() %>%
    st_as_sf()
}

# ── Entropy ───────────────────────────────────────────────────────────────────

# Normalized Shannon entropy for a vector of flows, scaled to [0, 1].
# n is the maximum possible number of distinct origins/destinations (total nodes
# minus one, since self-loops are excluded), which sets the scale.
entropy <- function(values, n = length(values)) {
  normed <- values / sum(values)
  # 0 * log(0) is defined as 0 in information theory; ifelse prevents NaN
  -sum(ifelse(normed == 0, 0, normed * log2(normed))) / log2(n)
}

stopifnot(
  entropy(c(5, 5, 5), 3) == 1, # perfectly uniform  -> max entropy
  entropy(c(0, 0, 100), 3) == 0, # single source      -> min entropy
  entropy(c(100), 3) == 0 # single value       -> min entropy
)

add_entropy <- function(g) {
  N <- vcount(g)
  g %>%
    activate(nodes) %>%
    mutate(
      in_entropy = map_dbl(1:N, \(node) {
        # In-flows to this node; max possible sources = N - 1 (no self-loops)
        g %>%
          activate(edges) %>%
          filter(to == node) %>%
          pull(n) %>%
          entropy(N - 1)
      }),
      out_entropy = map_dbl(1:N, \(node) {
        # Out-flows from this node; max possible destinations = N - 1
        g %>%
          activate(edges) %>%
          filter(from == node) %>%
          pull(n) %>%
          entropy(N - 1)
      })
    )
}

# ── Map helpers ───────────────────────────────────────────────────────────────

# Returns a function `year -> tmap` that fills CCAs by `variable`
map_by <- function(variable, legend_label, title, scale) {
  function(year) {
    get_sf(year) %>%
      tm_shape() +
      tm_polygons(
        fill        = variable,
        fill.legend = tm_legend(title = legend_label),
        fill.scale  = scale,
      ) +
      tm_title(glue("{title} ({year})"))
  }
}

# Diverging red-blue map showing percent change from 2002 to 2022.
# limits = c(-1, 1) uses a fixed ±100% scale with labelled breaks;
# limits = NULL lets tmap auto-fit the scale to the data.
pct_change_map <- function(df, legend_title, limits = c(-1, 1)) {
  fill_scale <- if (is.null(limits)) {
    tm_scale_continuous(values = "brewer.rd_bu", midpoint = 0)
  } else {
    tm_scale_continuous(
      values         = "brewer.rd_bu",
      limits         = limits,
      labels         = c("-100%", "-50%", "0%", "+50%", "+100% or more"),
      midpoint       = 0,
      outliers.trunc = c(TRUE, TRUE)
    )
  }
  df %>%
    tm_shape() +
    tm_polygons(
      fill        = "pct_change",
      fill.scale  = fill_scale,
      fill.legend = tm_legend(title = legend_title)
    )
}

# Diverging red-blue map showing raw (absolute) change from 2002 to 2022
raw_change_map <- function(df, legend_title) {
  df %>%
    tm_shape() +
    tm_polygons(
      fill = "change",
      fill.scale = tm_scale_continuous(
        values   = "brewer.rd_bu",
        limits   = c(-0.2, 0.2),
        labels   = c("-0.2", "-0.1", "0", "+0.1", "+0.2"),
        midpoint = 0
      ),
      fill.legend = tm_legend(title = legend_title)
    )
}

# Returns a data frame of flow-weighted mean commute distance per CCA for `year`
get_mean_distances <- function(year) {
  g(year) %>%
    activate(edges) %>%
    as_tibble() %>%
    # cca_distances is keyed by CCA name, not by the graph node indices
    left_join(cca_distances, by = c("from_name" = "from", "to_name" = "to")) %>%
    group_by(from_name) %>%
    summarize(mean_distance = sum(n * distance) / sum(n), .groups = "drop") %>%
    rename(name = from_name)
}

map_mean_distance <- function(year) {
  g(year) %>%
    activate(nodes) %>%
    left_join(get_mean_distances(year), by = "name") %>%
    as_tibble() %>%
    st_as_sf() %>%
    tm_shape() +
    tm_polygons(
      fill        = "mean_distance",
      fill.legend = tm_legend(title = "Mean distance (m)"),
      fill.scale  = tm_scale_continuous(values = "viridis", limits = c(4000, 25000))
    ) +
    tm_title(glue("Mean commute distance ({year})"))
}
```

```{r data-prep, include=FALSE}
g2002 <- create_graph(cca_flows_2002) %>% add_entropy()
g2022 <- create_graph(cca_flows_2022) %>% add_entropy()

map_in_degree <- map_by("in_degree", "In-commuting flow", "In-degree", tm_scale_continuous_log(values = "brewer.oranges", limits = c(100, 300000)))
map_out_degree <- map_by("out_degree", "Out-commuting flow", "Out-degree", tm_scale_continuous_log(values = "brewer.blues", limits = c(500, 40000)))
map_eigen <- map_by("eigen", "Eigenvector centrality", "Eigenvector Centrality", tm_scale_continuous_log(values = "viridis", limits = c(0.0001, 1)))
map_in_entropy <- map_by("in_entropy", "In-entropy", "In-entropy", tm_scale_continuous_log(values = "brewer.reds", limits = c(0.75, 1)))
map_out_entropy <- map_by("out_entropy", "Out-entropy", "Out-entropy", tm_scale_continuous_log(values = "brewer.purples", limits = c(0.4, 0.9)))

vars <- c("in_degree", "out_degree", "eigen", "in_entropy", "out_entropy")

# Geometry is taken from 2022; 2002 contributes only the metric columns
combined <- left_join(
  g2002 %>% activate(nodes) %>% as_tibble() %>% select(name, all_of(vars)),
  get_sf(2022) %>% select(name, all_of(vars)),
  by     = "name",
  suffix = c("_2002", "_2022")
) %>% st_as_sf()
```

## Network overview

The 2022 network contains `r vcount(g2022)` nodes and
`r g2022 %>% activate(edges) %>% filter(n > 0) %>% ecount()` non-zero directed
edges.

## Commuter origins

Not all LODES workers who work in Chicago live there; this table breaks down
2022 tract-level flows by whether the home and work tracts are inside city
limits.

```{r commuter-origins}
tract_to_cca <- tracts %>%
  as_tibble() %>%
  select(GEOID, cca)

# cca_flows only covers intra-Chicago flows; re-join from raw tract data to
# capture workers commuting in from outside the city
read_csv("data/tract_flows_2022.csv", col_types = "cci") %>%
  left_join(tract_to_cca, by = join_by(w_tract == GEOID)) %>%
  rename(w_cca = cca) %>%
  left_join(tract_to_cca, by = join_by(h_tract == GEOID)) %>%
  rename(h_cca = cca) %>%
  mutate(
    works_in_chicago = !is.na(w_cca),
    lives_in_chicago = !is.na(h_cca)
  ) %>%
  group_by(works_in_chicago, lives_in_chicago) %>%
  summarize(n = sum(n), .groups = "drop") %>%
  kable(
    col.names   = c("Works in Chicago", "Lives in Chicago", "Workers"),
    format.args = list(big.mark = ",")
  )
```

## Node–link diagram

Edges are filtered to the top 40% of flows; nodes are colored by distance
from the Loop.

```{r network-plot, fig.width=12, fig.height=8}
set.seed(1)
g2022 %>%
  activate(edges) %>%
  filter(n > quantile(n, 0.6)) %>%
  activate(nodes) %>%
  mutate(distance_from_loop = distance_from_loop / 1000) %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = n), edge_width = 0.2, edge_color = "#000") +
  geom_node_point(aes(color = distance_from_loop), size = 3) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  theme_graph(plot_margin = margin(10, 10, 10, 10)) +
  theme(
    legend.background      = element_rect(),
    legend.margin          = margin(8, 8, 8, 8),
    legend.position        = "inside",
    legend.position.inside = c(0.95, 0.05),
    legend.justification   = c(1, 0),
    legend.box.just        = "right",
    legend.box.margin      = margin(6, 6, 6, 6),
  ) +
  labs(color = "Distance from the Loop (km)", edge_alpha = "Commuting flow")
```

## Centrality

### In-degree

Total weighted in-commuting flow to each CCA — how many workers commute *into*
each neighborhood from another CCA.

```{r in-degree}
map_in_degree(2002)
map_in_degree(2022)
combined %>%
  mutate(pct_change = in_degree_2022 / in_degree_2002 - 1) %>%
  pct_change_map("In-degree % change")
```

```{r in-degree-scatter, eval=FALSE, include=FALSE}
# Scatter plot comparing 2002 vs. 2022 in-degree (kept for reference)
combined %>%
  ggplot(aes(x = in_degree_2002, y = in_degree_2022)) +
  geom_point() +
  geom_text_repel(aes(label = name)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()
```

### Out-degree

Total weighted out-commuting flow from each CCA — how many workers leave each
neighborhood to work somewhere else.

```{r out-degree}
map_out_degree(2002)
map_out_degree(2022)
combined %>%
  mutate(pct_change = out_degree_2022 / out_degree_2002 - 1) %>%
  pct_change_map("Out-degree % change")
```

### Eigenvector centrality

Eigenvector centrality weights a CCA's importance by the importance of its
neighbors, capturing which areas are connected to other highly-connected areas.

```{r eigen}
map_eigen(2002)
map_eigen(2022)
combined %>%
  mutate(pct_change = eigen_2022 / eigen_2002 - 1) %>%
  pct_change_map("Eigenvector centrality % change")
```

<!-- Kept for reference; references undefined cca_graph and outdated variable names

```{r eigen-vs-distance, eval=FALSE, include=FALSE}
cca_graph %N>%
  as_tibble() %>%
  mutate(area = if_else(
    st_centroid(geometry) %>% st_coordinates() %>% .[, 2] >=
      st_centroid(filter(ccas, name == "Near West Side")$geometry) %>%
        st_coordinates() %>%
        .[, 2],
    "Northern",
    "Southern"
  )) %>%
  arrange(desc(eig_centrality)) %>%
  ggplot(aes(x = distance_from_loop, y = eig_centrality)) +
  geom_point(aes(color = area)) +
  geom_smooth(se = FALSE, method = "lm", size = 0.5, color = "black", formula = y ~ x) +
  geom_text_repel(aes(label = name), size = 2) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set1") +
  theme_classic() +
  labs(
    title    = "Eigenvector Centrality vs. Distance from the Loop",
    subtitle = '"Northern" neighborhoods are those at least as far north as the Near West Side.',
    x        = "Distance from the Loop (m)",
    y        = "Eigenvector Centrality",
    color    = "Area"
  )
```

-->

## Entropy

### In-entropy

In-entropy measures how evenly a CCA's in-commuting is spread across source
neighborhoods; a value of 1 means workers arrive in equal numbers from all
other CCAs.

```{r in-entropy}
map_in_entropy(2002)
map_in_entropy(2022)
combined %>%
  mutate(change = in_entropy_2022 - in_entropy_2002) %>%
  raw_change_map("In-entropy change")
```

### Out-entropy

Out-entropy measures how evenly a CCA's out-commuting is spread across
destination neighborhoods.

```{r out-entropy}
map_out_entropy(2002)
map_out_entropy(2022)
combined %>%
  mutate(change = out_entropy_2022 - out_entropy_2002) %>%
  raw_change_map("Out-entropy change")
```

## Summary statistics

Descriptive statistics for each metric across all 77 CCAs, plus paired t-tests
comparing 2002 and 2022.

```{r summary-stats}
var_labels <- c(
  in_degree   = "In-degree",
  out_degree  = "Out-degree",
  eigen       = "Eigenvector centrality",
  in_entropy  = "In-entropy",
  out_entropy = "Out-entropy"
)

# Descriptive statistics by variable and year
combined %>%
  as_tibble() %>%
  select(name, contains("_20")) %>%
  pivot_longer(
    contains("_20"),
    names_to      = c("variable", "year"),
    values_to     = "value",
    names_pattern = r"(^([a-z_]+)_(\d{4})$)"
  ) %>%
  mutate(variable = factor(variable, levels = names(var_labels), labels = var_labels)) %>%
  group_by(Variable = variable, Year = year) %>%
  summarize(
    Mean = mean(value),
    SD = sd(value),
    SE = SD / sqrt(n()),
    Median = median(value),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Descriptive statistics by variable and year")

# Paired t-tests: 2022 vs. 2002
map_dfr(names(var_labels), function(v) {
  tt <- t.test(combined[[glue("{v}_2022")]], combined[[glue("{v}_2002")]], paired = TRUE)
  tibble(
    Variable      = var_labels[v],
    `Mean 2002`   = mean(combined[[glue("{v}_2002")]]),
    `Mean 2022`   = mean(combined[[glue("{v}_2022")]]),
    `Mean change` = unname(tt$estimate),
    `t`           = unname(tt$statistic),
    `p-value`     = tt$p.value
  )
}) %>%
  kable(digits = 3, caption = "Paired t-tests: 2022 vs. 2002")
```

## Mean distance commuted

Flow-weighted mean commute distance for residents of each CCA, highlighting
which neighborhoods send workers farthest across the city.

```{r mean-distance}
map_mean_distance(2002)
map_mean_distance(2022)
get_sf(2022) %>%
  select(name) %>%
  left_join(get_mean_distances(2002) %>% rename(dist_2002 = mean_distance), by = "name") %>%
  left_join(get_mean_distances(2022) %>% rename(dist_2022 = mean_distance), by = "name") %>%
  mutate(pct_change = dist_2022 / dist_2002 - 1) %>%
  pct_change_map("Mean distance % change", limits = NULL)
```

## K-cores

K-core decomposition of the 2002 network, restricted to the top quartile of
flows, identifies the dense commuting spine at the city's core.

```{r k-cores}
g2002 %>%
  activate(edges) %>%
  filter(n > quantile(n, 0.75)) %>%
  activate(nodes) %>%
  mutate(coreness = node_coreness()) %>%
  as_tibble() %>%
  st_as_sf() %>%
  tm_shape() +
  tm_polygons(fill = "coreness", fill.scale = tm_scale_ordinal())
```

## Structural equivalence

CCAs are grouped by Pearson correlation of their combined in- and
out-commuting profiles; neighborhoods in the same cluster occupy similar
structural positions in the network regardless of whether they are adjacent.

```{r structural-equiv}
get_dend <- function(g) {
  adj <- as_adjacency_matrix(g, attr = "n", sparse = FALSE)
  # Stack rows and transposed columns so clustering reflects both in- and
  # out-commuting simultaneously
  mtx_both <- rbind(adj, t(adj))
  hclust(as.dist(1 - cor(mtx_both), upper = FALSE), method = "ward.D")
}

add_structural_equiv <- function(g) {
  if ("cluster" %in% colnames(g %>% activate(nodes) %>% as_tibble())) {
    g <- g %>%
      activate(nodes) %>%
      select(-cluster)
  }
  cluster_assignments <- cutree(get_dend(g), k = 5) %>%
    tibble(cluster = ., name = names(.))
  g %>%
    activate(nodes) %>%
    left_join(cluster_assignments, by = "name")
}

g2002 <- add_structural_equiv(g2002)
g2022 <- add_structural_equiv(g2022)

map_struct_equiv <- map_by(
  "cluster", "Structural equivalence class",
  "Structural Equivalence Clusters",
  scale = tm_scale_categorical()
)
tmap_arrange(map_struct_equiv(2002), map_struct_equiv(2022))

plot(get_dend(g2002))
plot(get_dend(g2022))
```

## Regular equivalence

Regular equivalence (REGE) groups nodes whose neighbors are similarly
structured, whether or not those nodes are directly connected to each other.

```{r regular-equiv}
mat <- as_adjacency_matrix(g2002, attr = "n", sparse = FALSE)
rege <- REGE.nm.for(mat)$E
dend <- hclust(as.dist(1 - rege), method = "ward.D")

plot(dend)
plot.mat(mat, clu = cutree(dend, k = 8))
```

<!-- =============================================================
     FEEDBACK NOTES (2025-04-17) — kept for reference, not rendered
     =============================================================

# Feedback 4/17/25

- Literature:
  - Check labor literature; Detroit
  - Redlining, history of Chicago
  - Gold Coast/Slum book. Great American City book.
  - Rob Simson (?)
- Issues with the regression idea:
  - Disruptions in 2008 and 2020
  - Would have to control for a lot of things: population, overall economic change
    (shouldn't be too hard)
- Misc
  - **How is virtual work captured in the survey?**
  - Compute degree without self-ties so what we're measuring is **cross-neighborhood
    commuting** patterns
  - Justify my choice of Chicago as the boundary
- Alternative ideas:
  - Just narrow to COVID time frame, see what changed in the data
  - Only do descriptive statistics. Longitudinal preferably. Could use blockmodeling
    or community detection techniques or structural/regular equivalence.

-->

<!-- =============================================================
     OLD IDEAS — kept for reference, not rendered
     =============================================================

# Old Ideas

## Where people who live in a given CCA work?

One simple result we can draw from this dataset is simply where people who live
in any given commuting area work. For example, below we compare the commuting
flows for residents of Hyde Park and Woodlawn. Hyde Park and Woodlawn are
adjacent, but because Hyde Park is home to the University of Chicago, most of
its residents work there. Woodlawn, on the other hand, has a shortage of local
jobs, so most Woodlawn residents commute to the Loop.

```{r where-people-work, eval=FALSE, include=FALSE}
where_do_people_work <- function(cca) {
  cca_flows_2022 %>%
    filter(from == cca) %>%
    select(to, n) %>%
    arrange(desc(n))
}

where_do_people_work_plot <- function(cca) {
  where_do_people_work(cca) %>%
    filter(to != "OUTSIDE CHICAGO") %>%
    mutate(is_source = to == cca) %>%
    arrange(is_source) %>%
    left_join(ccas, by = c("to" = "name")) %>%
    st_as_sf() %>%
    tm_shape() +
    tm_polygons(
      fill        = "n",
      fill.legend = tm_legend(title = "Number of Workers"),
      col         = "is_source",
      col.scale   = tm_scale_categorical(values = c(`TRUE` = "red", `FALSE` = "lightblue", `NA` = "lightblue")),
      col.legend  = tm_legend_hide(),
      lwd         = "is_source",
      lwd.scale   = tm_scale_categorical(values = c(`TRUE` = 2, `FALSE` = 0.5)),
      lwd.legend  = tm_legend_hide(),
    ) +
    tm_title(glue("Where do {cca} residents work?"), size = 1.5)
}

where_do_people_work_plot("North Lawndale")
```

## Local jobs availability

This plot shows the number of local jobs per working resident in each community
area. In neighborhoods with the largest employment, like O'Hare and the Loop,
there are many more jobs than working residents. By contrast, many
neighborhoods have nearly ten times as many working residents as jobs, so
almost everyone must commute.

```{r jobs-availability, eval=FALSE, include=FALSE}
ccas <- ccas %>%
  mutate(jobs_availability = w_total / h_total)

ccas %>%
  tm_shape() +
  tm_polygons(
    fill        = "jobs_availability",
    fill.legend = tm_legend(title = "Jobs Availability"),
    fill.scale  = tm_scale_continuous_log10(values = "viridis"),
  )
```

## Distance and flow size

Intuitively, we would expect commuting to be higher between neighborhoods that
are closer together.

```{r distance-vs-size, eval=FALSE, include=FALSE}
cca_flows_2022 %>%
  filter(from != "OUTSIDE CHICAGO", to != "OUTSIDE CHICAGO") %>%
  filter(n != 0) %>%
  ggplot(aes(x = distance, y = jitter(n, amount = 0.99))) +
  geom_point(size = 1, alpha = 0.3) +
  scale_y_log10() +
  theme_minimal() +
  guides(fill = "none") +
  labs(title = "Distance vs. Flow Size", x = "Distance (m)", y = "Flow Size")
```

## Modeling Commuting Patterns with an ERGM

Modeling this dataset with an exponential random graph model (ERGM) would
help understand what factors determine where people work. However, the ERGMs
we learned about in class only work with unweighted networks. Krivitsky (2012)
extended the ERGM framework to networks with values that represent counts,
which is what we have here.

```{r ergm-setup, eval=FALSE, include=FALSE}
chicago_only <- cca_graph %>%
  activate(nodes) %>%
  filter(name != "OUTSIDE CHICAGO")

net <- asNetwork(chicago_only)

net %v% "residents_log" <- log(chicago_only %>% activate(nodes) %>% pull(h_in_chicago))
net %v% "workers_log" <- log(chicago_only %>% activate(nodes) %>% pull(w_from_chicago))

# Full distance matrix; all dyads are represented in cca_flows
D <- cca_flows_2022 %>%
  filter(from != "OUTSIDE CHICAGO", to != "OUTSIDE CHICAGO") %>%
  select(from, to, distance) %>%
  pivot_wider(names_from = to, values_from = distance) %>%
  column_to_rownames("from") %>%
  as.matrix()
```

```{r ergm-1, eval=FALSE, include=FALSE}
fit1 <- ergm(
  net ~
    sum + # baseline intensity
    nonzero + # sparsity
    nodeocov("residents_log") + # origin size effect
    nodeicov("workers_log") + # destination size effect
    edgecov(D), # distance deterrence
  reference = ~Poisson,
  response = "n",
)
```

-->
